\documentclass[5p,authoryear]{elsarticle}

\journal{SoftwareX}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{orange},
  language=Python,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny,
}

\begin{document}

\begin{frontmatter}

\title{MOLEKUL: A Pure-Python Ab Initio Quantum Chemistry Platform for
       Education and Reproducible Benchmarking}

\author{Enis Yazici}
\ead{yazicienis@gmail.com}
\address{SRH University of Applied Sciences Heidelberg,
         School of Engineering and Architecture,
         Heidelberg, Germany}
\cortext[cor1]{Corresponding author}

\begin{abstract}
MOLEKUL is an open-source, pure-Python implementation of Restricted
Hartree--Fock (RHF) self-consistent field theory and second-order
M{\o}ller--Plesset perturbation theory (MP2) for closed-shell molecules.
Starting from an XYZ-format molecular geometry and a contracted Gaussian basis
set, MOLEKUL evaluates all one- and two-electron integrals analytically, solves
the Roothaan--Hall equations iteratively, and reports the total RHF and MP2
energies.
Built-in basis data cover hydrogen through fluorine for STO-3G, 6-31G*, and
cc-pVDZ.
Additional modules provide a geometry optimizer, Mulliken and L{\"o}wdin
population analysis, experimental excited-state infrastructure, and numerical
Hessian-based harmonic frequency workflows.
Validated against PySCF on a 14-molecule benchmark suite, MOLEKUL achieves
RHF/STO-3G total-energy agreement within $5\times10^{-8}$~hartree and MP2
correlation-energy agreement within $2\times10^{-7}$~hartree.
The codebase comprises 606 automated tests and is designed for advanced
undergraduate and graduate courses in quantum chemistry, as well as for
reproducible hardware benchmarking.
\end{abstract}

\begin{keyword}
quantum chemistry \sep
Hartree--Fock \sep
MP2 \sep
ab initio \sep
Gaussian basis sets \sep
Python \sep
education \sep
benchmarking
\end{keyword}

\end{frontmatter}

% ============================================================
\section*{Code Metadata}
% ============================================================

\begin{table}[h]
\caption{Code metadata}
\begin{tabular}{lp{0.55\textwidth}}
\toprule
\textbf{Nr.} & \textbf{Code metadata description} \\
\midrule
C1 & Current code version: v0.1.1 \\
C2 & Permanent link to code/repository:
     \url{https://github.com/yazicienis/molekul} \\
C3 & Permanent link to reproducible capsule (Zenodo):
     \url{https://doi.org/10.5281/zenodo.19743306} \\
C4 & Legal code license: MIT \\
C5 & Code versioning system: git \\
C6 & Software languages and tools: Python~$\ge$~3.10, NumPy \\
C7 & Dependencies: \texttt{numpy}; \texttt{pytest} (development) \\
C8 & Developer documentation:
     \url{https://github.com/yazicienis/molekul} \\
C9 & Support contact: \texttt{yazicienis@gmail.com} \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Motivation and Significance}
% ============================================================

Production electronic-structure packages such as PySCF~\citep{sun2020},
Psi4~\citep{smith2020}, and ORCA~\citep{neese2020} provide broad functionality
and optimized implementations, but their internal complexity can obscure the
connection between textbook equations and executable algorithms.
Earlier educational codes such as PyQuante~\citep{quante2004} demonstrated the
value of readable implementations, but they are no longer as actively maintained
or systematically validated against modern reference packages.

MOLEKUL occupies the space between textbook equations and production code.
Every algorithmic step---from primitive integral evaluation to Fock matrix
diagonalization---traces directly to a named Python function with clear
correspondence to standard quantum-chemistry references~\citep{szabo1989,
helgaker2000}.
The 606-test suite and deterministic NumPy implementation ensure that every
numerical claim is automatically reproducible.

Beyond pedagogy, the Python/NumPy implementation provides a self-contained
baseline for hardware benchmarking, because quantum-chemistry algorithms are
implemented explicitly rather than delegated to compiled electronic-structure
kernels, allowing clean isolation of NumPy/BLAS performance characteristics.

% ============================================================
\section{Software Description}
% ============================================================

\subsection{Architecture}

MOLEKUL is organized as an installable Python package (\texttt{src/molekul/}).
Table~\ref{tab:modules} lists the main modules.
The package is installed with \texttt{pip install -e ".[dev]"} and tested with
\texttt{pytest tests/}.

\begin{table}[h]
\caption{Main modules of MOLEKUL.}
\label{tab:modules}
\begin{tabular}{ll}
\toprule
\textbf{Module} & \textbf{Responsibility} \\
\midrule
\texttt{molecule.py}, \texttt{atoms.py}       & Molecular geometry and atomic data \\
\texttt{basis\_sto3g.py}, \texttt{basis\_631gstar.py}, \texttt{basis\_ccpvdz.py}
                                               & Contracted Gaussian basis sets \\
\texttt{integrals.py}                          & One-electron integrals (overlap, kinetic,
                                                 nuclear, dipole) \\
\texttt{eri.py}                                & Two-electron repulsion integrals \\
\texttt{rhf.py}                                & RHF SCF driver (DIIS, SAD guess) \\
\texttt{mp2.py}                                & MP2 correlation energy \\
\texttt{population.py}                         & Mulliken and L{\"o}wdin population analysis \\
\texttt{optimizer.py}                          & Geometry optimizer (numerical gradients) \\
\texttt{harmonic.py}                           & Harmonic frequency analysis \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Integral Engine}

One- and two-electron integrals are evaluated analytically using the
McMurchie--Davidson (MD) scheme~\citep{mcmurchie1978,helgaker2000}.
The MD scheme expresses products of Cartesian Gaussians as a linear combination
of Hermite Gaussians via expansion coefficients, reducing all integral
evaluation to a Boys-function call
$F_n(x) = \int_0^1 t^{2n} e^{-xt^2}\,\mathrm{d}t$~\citep{boys1950}.

The Boys function is evaluated using a pure-Python combination of
\texttt{math.erf} for the analytic $F_0$ expression, recurrence relations, and
small-$x$ Taylor expansions, with the branch selected to maintain numerical
stability over the argument range used in the validation suite.
No SciPy dependency is required; a dedicated test confirms relative errors below
$5\times10^{-14}$ for the $n$ and $x$ values encountered in STO-3G through
cc-pVDZ bases.

Two-electron repulsion integrals (ERIs) are stored in a full $N_\mathrm{AO}^4$
double-precision array.
Dense storage is intentional---it keeps the Fock-build loop transparent and
inspectable.
At $N_\mathrm{AO}=100$ this requires approximately 800~MB; practical use is
therefore limited to small molecules and teaching-scale basis sets.
Integral screening, density fitting, and integral-direct algorithms are
documented extension points.

Contracted Gaussian shells are explicitly renormalized to unit overlap,
correcting for cross-primitive overlap contributions commonly omitted in
introductory treatments.

\subsection{SCF Driver}

The RHF SCF driver~\citep{roothaan1951,hall1951} implements three convergence
features:
\begin{itemize}
  \item \textbf{SAD initial guess}: a Superposition of Atomic Densities density
        matrix with spherically-averaged neutral-atom occupations, avoiding
        wrong-state convergence traps from the core-Hamiltonian guess.
  \item \textbf{DIIS acceleration}: Pulay's Direct Inversion in the Iterative
        Subspace~\citep{pulay1980,pulay1982} using the error vector
        $\mathbf{e}=\mathbf{FPS}-\mathbf{SPF}$.
  \item \textbf{Level shift}: an optional virtual-orbital energy
        shift~\citep{saunders1973} that stabilizes early iterations and is
        disabled once the DIIS subspace is populated.
\end{itemize}
Default convergence thresholds: $|\Delta E|<10^{-10}$~E$_h$ and
$\max|\Delta\mathbf{P}|<10^{-8}$.

\subsection{Post-HF and Analysis}

The MP2~\citep{moller1934} correlation energy is computed from the converged RHF
molecular orbitals via a four-index integral transformation.
Mulliken and L{\"o}wdin~\citep{lowdin1950} population analysis, electric dipole
moments, and a geometry optimizer based on numerical nuclear gradients are also
implemented.
Excited-state and harmonic frequency analysis modules are available as
experimental pedagogical infrastructure.

\subsection{Basis Sets}

Built-in basis data cover hydrogen through fluorine.
Exponents and contraction coefficients are taken directly from PySCF's
built-in basis-set data~\citep{pritchard2019} to ensure reproducibility.

\begin{table}[h]
\caption{Built-in basis sets.}
\label{tab:basis}
\begin{tabular}{lll}
\toprule
\textbf{Basis} & \textbf{Reference} & \textbf{AO per heavy atom} \\
\midrule
STO-3G  & \citet{hehre1969}                        & 5  \\
6-31G*  & \citet{hehre1972}; \citet{hariharan1973} & 15 \\
cc-pVDZ & \citet{dunning1989}                      & 15 \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Illustrative Example}
% ============================================================

The following example computes the RHF and MP2 energies of water at a standard
STO-3G geometry:

\begin{lstlisting}[language=Python, caption={RHF and MP2 energies of H$_2$O.}]
from molekul.molecule import Molecule
from molekul.atoms import Atom
from molekul.basis_sto3g import get_sto3g
from molekul.rhf import rhf_scf
from molekul.mp2 import mp2_energy
import numpy as np

ANG2BOHR = 1.8897259886

mol = Molecule([
    Atom('O', np.array([ 0.000,  0.000,  0.117]) * ANG2BOHR),
    Atom('H', np.array([ 0.000,  0.757, -0.469]) * ANG2BOHR),
    Atom('H', np.array([ 0.000, -0.757, -0.469]) * ANG2BOHR),
])

basis = get_sto3g()
rhf_result  = rhf_scf(mol, basis)
print(f"RHF energy: {rhf_result.energy_total:.8f} Eh")  # -74.96258854 Eh

mp2_result  = mp2_energy(mol, basis, rhf_result)
print(f"MP2 energy: {mp2_result.energy_total:.8f} Eh")  # -74.99844967 Eh
\end{lstlisting}

% ============================================================
\section{Validation}
% ============================================================

\subsection{RHF/STO-3G Benchmark}

The benchmark script \texttt{scripts/benchmark\_14mol.py} compares MOLEKUL
RHF/STO-3G total energies against PySCF~\citep{sun2020} for 14 closed-shell
molecules spanning H$_2$ through formaldehyde (CH$_2$O).
All 14 molecules pass with maximum absolute deviation $4.9\times10^{-8}$~E$_h$
(C$_2$H$_2$), consistent with floating-point differences in integral
evaluation.
Complete results are logged in \texttt{outputs/logs/benchmark\_14mol.json}.

\subsection{MP2 Correlation Energy}

Table~\ref{tab:mp2} compares MP2 correlation energies against PySCF for eight
STO-3G molecules spanning polar hydrides, homonuclear diatomics, and polyatomic
systems.
All eight molecules agree within $2\times10^{-7}$~E$_h$.

\begin{table}[h]
\caption{MP2 correlation energies: MOLEKUL vs.\ PySCF (STO-3G).}
\label{tab:mp2}
\begin{tabular}{lrr}
\toprule
\textbf{Molecule} &
\textbf{$E_\mathrm{corr}^\mathrm{PySCF}$ / E$_h$} &
\textbf{$|\Delta E_\mathrm{corr}|$ / E$_h$} \\
\midrule
H$_2$  & $-0.01313807$ & $8.1\times10^{-10}$ \\
HF     & $-0.01734432$ & $2.6\times10^{-9}$  \\
H$_2$O & $-0.03550283$ & $1.4\times10^{-8}$  \\
N$_2$  & $-0.15419856$ & $2.2\times10^{-8}$  \\
CO     & $-0.12852242$ & $1.2\times10^{-7}$  \\
NH$_3$ & $-0.04704176$ & $5.6\times10^{-9}$  \\
CH$_4$ & $-0.05650741$ & $6.1\times10^{-9}$  \\
HCN    & $-0.12997898$ & $5.4\times10^{-9}$  \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Known Limitations}
% ============================================================

\begin{itemize}
  \item \textbf{Memory}: dense $N_\mathrm{AO}^4$ ERI storage limits practical
        use to small molecules ($N_\mathrm{AO}\lesssim50$).
  \item \textbf{Single-threaded}: no explicit parallelism; single NumPy/BLAS
        thread by default.
  \item \textbf{Closed-shell RHF only}: UHF and ROHF are not supported.
  \item \textbf{Element coverage}: H--F only; heavier elements are not
        parametrized.
  \item \textbf{Numerical derivatives}: geometry optimization and
        harmonic-frequency workflows use finite differences, not analytic
        gradients.
  \item \textbf{Experimental modules}: excited-state and frequency-analysis
        workflows are pedagogical infrastructure, not production-validated.
  \item No integral screening, effective core potentials, or relativistic
        corrections.
\end{itemize}

% ============================================================
\section{Impact}
% ============================================================

MOLEKUL serves two complementary purposes.
As a teaching tool, all algorithms trace to named functions in readable Python,
making it suitable for advanced undergraduate and graduate courses in quantum
chemistry and computational physics.
As a reproducible benchmarking baseline, the 606-test suite and deterministic
NumPy~\citep{harris2020} implementation allow hardware and software comparisons
without delegating quantum-chemistry algorithms to compiled electronic-structure
kernels.
Benchmark timings and energies are logged to version-controlled JSON files for
full reproducibility.

% ============================================================
\section*{Software Availability}
% ============================================================

\begin{itemize}
  \item \textbf{Repository:} \url{https://github.com/yazicienis/molekul}
  \item \textbf{Archive DOI:} \url{https://doi.org/10.5281/zenodo.19743306}
  \item \textbf{License:} MIT
  \item \textbf{Version:} v0.1.1
  \item \textbf{Installation:} \texttt{pip install -e ".[dev]"}
  \item \textbf{Tests:} \texttt{pytest tests/}
\end{itemize}

% ============================================================
\section*{AI Usage Disclosure}
% ============================================================

Generative AI tools, including Claude (Anthropic) and ChatGPT (OpenAI), were
used during software development for code drafting, refactoring assistance,
documentation support, test-generation suggestions, and editorial review of
the manuscript.
All scientific algorithms, numerical reference values, validation criteria, and
reported results were reviewed and verified by the author, who made the primary
design decisions and remains responsible for the correctness, originality, and
scientific content of the submission.
Reference values were not modified to fit the implementation.

% ============================================================
\section*{Acknowledgements}
% ============================================================

The author thanks the developers of PySCF, whose reference calculations were
used throughout validation, and the EMSL Basis Set Exchange~\citep{pritchard2019}
for tabulated basis-set parameters.

% ============================================================
\bibliographystyle{elsarticle-harv}
\bibliography{paper}
% ============================================================

\end{document}
